Recurrent Deep Learning Models and its applications

### 2025-02-09 paper2/main.tex
paper update

\newpage
\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general}. In rigorous terms, let the state space be defined as:
Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

\newpage
\textbf{Hidden markov model}
\\[1ex]
Hidden Markov Model (HMMs) is a statistical model that extends the Markov chains by introducting a layer of latent (unobservable) variables. And connecting them to observable outputs (emission probability). In contrast to standard Markov chains, hidden states are never be observed directly. Instead, we observe a sequence of data (observations) that are probabilitically generated by these hidden state. This structure has proven useful for applications in speech recognition, biological sequence analysis, and signal processing \parencite{rabiner1989atutorial}. Lets defined the 
% ####################################################
\newpage

Literature Review on Hidden Markov Models

Hidden Markov Models (HMMs) extend classical Markov chains by incorporating an additional layer of latent (unobservable) variables that govern the generation of the observed data. In contrast to standard Markov chains—where the state of the system is directly observed—an HMM assumes that the underlying process is hidden and that each observed datum is stochastically generated from the corresponding hidden state. This two-layer structure has proven to be both theoretically interesting and practically useful for applications in speech recognition, biological sequence analysis, and signal processing (Rabiner, 1989).

1. From Markov Chains to Hidden Markov Models

Let \(\{X_n\}_{n \ge 1}\) be a Markov chain on a measurable space \((\mathcal{X}, \mathcal{F})\) satisfying
\[
P\bigl(X_{n+1} \in A \mid X_n, X_{n-1}, \dots, X_1\bigr) = P\bigl(X_{n+1} \in A \mid X_n\bigr) \quad \text{for all } A \in \mathcal{F}.
\]
This *Markov property* ensures that the future evolution of the process depends only on its present state, not on the full past history.

In extending this framework to HMMs, we introduce an observable process \(\{Y_n\}_{n \ge 1}\) defined on another measurable space \((\mathcal{Y}, \mathcal{G})\). The connection between the hidden states and the observations is governed by the following assumption:

- **Conditional Independence of Observations:** Given the current hidden state,
  \[
  P\bigl(Y_n \in B \mid X_n, X_{n-1}, \dots, X_1, Y_{n-1}, \dots, Y_1\bigr) = P\bigl(Y_n \in B \mid X_n\bigr)
  \]
  for every \(B \in \mathcal{G}\). This states that the observation \(Y_n\) is independent of past states and observations, conditioned on \(X_n\).

Thus, an HMM is fully specified by the following three components:
1. The **initial distribution** \(\pi\) on \(\mathcal{X}\):
   \[
   \pi(x) = P(X_1 = x).
   \]
2. The **transition kernel** \(P(x, dy)\) of the hidden Markov chain:
   \[
   p_{ij} = P(X_{n+1} = j \mid X_n = i).
   \]
3. The **emission (observation) distribution** \(B(x, dy)\) or \(b_x(y)\):
   \[
   b_x(y) = P(Y_n = y \mid X_n = x).
   \]

2. Joint and Marginal Distributions

Using the above structure, the joint probability of a sequence of hidden states and observations \(\{(X_1, Y_1), \dots, (X_N, Y_N)\}\) is given by
\[
\begin{aligned}
P\bigl(X_1, \dots, X_N,\, Y_1, \dots, Y_N\bigr) 
&= \pi(X_1) \, b_{X_1}(Y_1) \prod_{n=2}^N P\bigl(X_n \mid X_{n-1}\bigr) \, b_{X_n}(Y_n)\\[1mm]
&= \pi(X_1)\, b_{X_1}(Y_1)\, p_{X_1X_2}\, b_{X_2}(Y_2) \cdots p_{X_{N-1}X_N}\, b_{X_N}(Y_N).
\end{aligned}
\]
Marginalization over the hidden states yields the likelihood of the observations:
\[
P(Y_1, \dots, Y_N) = \sum_{X_1 \in \mathcal{X}} \cdots \sum_{X_N \in \mathcal{X}} \pi(X_1)\, \prod_{n=2}^N p_{X_{n-1}X_n}\, \prod_{n=1}^N b_{X_n}(Y_n),
\]
which in many applications is used for parameter estimation and inference.

3. Inference and Algorithmic Developments

Due to the latent structure in HMMs, direct computation of the likelihood or the posterior distribution of the hidden states is usually intractable when the state space is large or continuous. This has spurred the development of several efficient algorithms:

- **The Forward-Backward Algorithm:** Utilizes dynamic programming to recursively compute the probabilities of partial observation sequences. For instance, denoting the forward variable by
  \[
  \alpha_n(x) = P(Y_1, Y_2, \dots, Y_n, X_n = x),
  \]
  one can compute
  \[
  \alpha_{n+1}(y) = \left[\sum_{x \in \mathcal{X}} \alpha_n(x) \, p_{xy}\right]\, b_y(Y_{n+1}).
  \]

- **The Viterbi Algorithm:** Determines the most probable sequence of hidden states (i.e., finds \(\hat{X}_1, \dots, \hat{X}_N\) such that the joint probability is maximized).

- **The Baum-Welch Algorithm (Expectation-Maximization):** Iteratively estimates the unknown parameters (transition probabilities and emission probabilities) of the HMM when they are not known a priori.

These methods rely on the underlying assumption that the hidden state process is a Markov chain, and many of the rigorous convergence and mixing properties discussed in the literature on general state space Markov chains (e.g., Roberts and Rosenthal, 2004; Meyn and Tweedie, 1993) carry over—often with additional considerations due to the unobserved structure.

4. Theoretical Underpinnings and Convergence

As with observable Markov chains, the study of HMMs involves rigorous mathematical analysis centered on convergence properties and ergodicity. Under appropriate conditions (e.g., the hidden chain is \(\varphi\)-irreducible, aperiodic, and satisfies a suitable drift condition), one can prove that the filtering distributions
\[
P(X_n \in \cdot \mid Y_1, \dots, Y_n)
\]
converge to a unique invariant measure. Techniques such as **coupling** and **regeneration**—which have been successfully employed in establishing quantitative convergence rates for Markov chains (Roberts \& Rosenthal, 2004)—are also adapted to the HMM context to study the stability and consistency of inference procedures.

For example, under the standard assumptions of an HMM, if the hidden state chain is geometrically ergodic then for a function \(h : \mathcal{X} \to \mathbb{R}\) with finite second moment under the stationary measure \(\pi\), the central limit theorem holds for the additive functional
\[
\frac{1}{\sqrt{n}} \sum_{i=1}^n \left[h(X_i) - \pi(h)\right],
\]
which is essential for quantifying the uncertainty in estimators derived via the Expectation-Maximization or sequential Monte Carlo methods (see also Billingsley, 1995).

5. Conclusion

In summary, by extending the framework of Markov chains—characterized by the Markov property, established convergence criteria, small set theory, and drift conditions—to a hidden state setting, Hidden Markov Models provide a robust foundation for dealing with complex sequence data where the underlying process is not directly observable. This extension leads to rich algorithmic and inferential techniques that have been rigorously analyzed and widely adopted in several applied fields.
% ######################## Background and Literature Review #########################

>>>>

\newpage
\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general, rabiner1986anintroduction}. In rigorous terms. Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

@@

% Hidden markov model 
\textbf{Hidden markov model}
\\[1ex]
Hidden Markov Model (HMMs) is a statistical model that extends the Markov chains by introducting a layer of latent (unobservable) variables. And connecting them to observable outputs (emission probability). In contrast to standard Markov chains, hidden states are never be observed directly. Instead, we observe a sequence of data (observations) that are probabilitically generated by these hidden state. This structure has proven useful for applications in speech recognition, biological sequence analysis, and signal processing \parencite{rabiner1989atutorial}. Lets the hidden states be denoted by $X_t$, where $X_t$ is the hidden state at time $t$. The hidden state space is defined as:
\begin{equation}
    \mathcal{X} = \{s_1, s_2, \cdots, s_n\}
\end{equation}
where $\mathcal{X}$ is finite set of size $n$, and each $s_i \in \mathcal{X}$ represents a specific possible hidden state. Then for the observable space represents the possible outcomes (observations) that are generated by the hidden state, we have discuss previously. Let the observations at time $t$ be denoted by $Y_t$. The observable space is defined as:
\begin{equation}
    \mathcal{Y} = \{o_1, o_2, \cdots, o_m\}
\end{equation}
where $\mathcal{Y}$ is finite set of size $m$, and each $o_i \in \mathcal{Y}$ represents a specific possible observation. 
\\[1ex]
Then for the transition probability matrix ($P$), the probability of transitioning between hidden states denoted as:
\begin{equation}
    P_{n \times n} = 
    \left[ {\begin{array}{cccc}
        p_{11} & p_{12} & \cdots & p_{1n}\\
        p_{21} & p_{22} & \cdots & p_{2n}\\
        \vdots & \vdots & \ddots & \vdots\\
        p_{n1} & p_{n2} & \cdots & p_{nn}\\
    \end{array} } \right]
\end{equation}
\begin{equation}
    p_{ij} = P(X_{t+1} = s_j \mid X_t = s_i) 
\end{equation}
where $s_j, s_i \in \mathcal{X}$.
\\[1ex]
For emission probability matrix ($B$), the probabilities of generating an observation given a hidden state.
\begin{equation}
    B_{n \times n} = 
    \left[ {\begin{array}{cccc}
        b_{11} & b_{12} & \cdots & b_{1m}\\
        b_{21} & b_{22} & \cdots & b_{2m}\\
        \vdots & \vdots & \ddots & \vdots\\
        b_{m1} & b_{m2} & \cdots & b_{mm}\\
    \end{array} } \right]
\end{equation}
\begin{equation}
    b_{ik} = P(Y_{t} = o_k \mid X_t = s_i) 
\end{equation}
where $s_i \in \mathcal{X}, o_k \in \mathcal{Y}$.
\\[1ex]
And the goal of hidden markov model is the maximize the joint probability $P(X, Y)$ given the initial probability $P(X_1)$.
\begin{equation}
    P(X, Y) = P(X_1) \cdot \prod_{t=1}^{T-1}P(X_{t+1}=s_j \mid X_{t}=s_i) \cdot \prod_{t=1}^{T} P(Y_t=o_k \mid X_t=s_i)
\end{equation}

For the NLP tasks, there may have some unobservable data, for example the implicit sentiment, user intent, emotional state. \parencite{rabiner1993fundamentals} proposes the hidden Markov model (HMM), the assumption is the existence of the latent process follows a Markov chain from which observations X are generated. In other word, there would exists an unobserved state sequence $Z = \{z_1, z_2 \cdots , z_T\}$ in observed sequence $X = \{x_1, x_2, \cdots, x_T\}$ \parencite{sengupta2023hybrid}. Where the hidden states, $z_t$ belonging to state-space $Q = \{q_1, q_2, \cdots, q_M \}$ follow a Markov chain goverened by: 
\begin{itemize}
    \item A state-transition probability matrix $A = [a_{ij}] \in \mathbb{R}^{M \times M}$ where $a_{ij} = p(z_{t+1} = q_j | z_t = q_i)$ 
\end{itemize}
\begin{itemize}
    \item Initial state matrix $\pi = [\pi_i] \in \mathbb{R}_{1 \times M}$ with $\pi_i = p(z_1 = q_i)$
\end{itemize}
Furthermore, for the hidden state $z_t$, correspoinding to the observe data $x_t$ is release by emission process $B = [b_j(x)]$ where $b_j(x) = p(x|z=q_i)$. We can assume $b_j(x)$ is follows the Gaussian mixture model (GMM).
\begin{equation}
    p(x_t|z=q_j) = \sum_{l=1}^{k}c_{jl}\mathcal{N}(x_t|\mu_{jl}, \Sigma_{jl})
\end{equation}
where $\sum_{l=1}^{k}c_{jl} = 1$, $\forall j = \{1,\cdots, M\}$, k is the number of Gaussian mixture components and $\mathcal{N}(x_t|\mu_{jl}, \Sigma_{jl})$ denotes a Gaussian probability density with mean $\mu_{jl}$ and covariance $\Sigma_{jl}$ for state $j$ and mixture component $l$. The number of hidden states ($M$) and mixture component ($k$) are the two hyperparameters of the model which have to be provided apriori.
\\[1ex]
Therefore, the joint probability probability density function of the observation $X$ can be expressed as:
\begin{equation}
    p(X) = p(z_1) \prod_{t=1}^{T-1} p(z_{t+1}|z_{t}) \prod_{t=1}^{T} p(x_t|z_t)
\end{equation}
The optimal parameters $[A, B, \pi]$, which maximize the likelihood of the observation sequence $X$ (Equation 39), are determined using the Baum-Welch algorithm, an expectation-maximization method \parencite{rabiner1993fundamentals}. Additionally, the probability of the system in a specific hidden state $z_t$ corresponding to the observation $x_t$ is calculated using the Viterbi algorithm.

% Word representation 
\subsubsection{Word representation}
In RNN, the due to the architecture, the model can only handle vector, it is important that convert the word into machine readable format. By translating words into vectors, these representations capture semantic meanings, relationships and contexts. \parencite{mikolov2013efficient} proposed 2 model architectures for leaning distributed representations of words.
\\[1ex]
\textbf{Continuous Bag of Words model}
CBOW operates by taking context words around a target word, it is flexible to adjust the window size which mean that we can control how many words around our target word. Hence aggregating their embeddings and passing to hidden layer to produce a probability distribution. Capture the semantic meanings and the relationship more effectively. 
\begin{equation}
    Q = N \times D + D \times log_2(V)
\end{equation}
\\[1ex]
\textbf{Continuous Skip gram model}
\\[1ex]
The continuous skip gram model operates in opposite direction compare with CBOW. The model takes the surrounding word to predict the target word. In other word, treat the current word as an input to a log-linear classifier with continuous projection layer, and predict the probability distribution of surrounding words.
\begin{equation}
    Q = C \times (D + D \times log_2(V))
\end{equation}
\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\textwidth]{../Pic/word_representation.png} % Replace "example-image" with your image file
\end{figure}

% ######################## Reference ######################################
\printbibliography
% ######################## Reference ######################################
